---
title: "R Project Milestone 6"
author: "Moyra Rasheed, Courtney Coon, Jarett Maycott"
date: '2022-11-28'
output: html_document

---


```{r setup, include=FALSE}
# BACKGROUND: packages
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(kableExtra)
```

```{r echo=FALSE, results=FALSE}
# BACKGROUND: Demo Data
## read in data
demo_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_county_demographic.csv",
                    stringsAsFactors = FALSE)

## subset demographic dataset to include only needed columns
demo_sub<-select(demo_data, c("pop2012","pop12_sqmi", "name", "med_age","renter_occ", "owner_occ"))

## categorize median age: younger is lower priority, older is high priority
sum_med_age<-summary(demo_sub$med_age)
#sum_med_age
demo_sub<-demo_sub%>%
  mutate(med_age_CAT= case_when(
    med_age < as.numeric(sum_med_age[2]) ~ "Low priority",      
    med_age < as.numeric(sum_med_age[5]) ~ "Medium priority",
    TRUE ~ "High priority"))%>% 
  mutate(med_age_CAT = factor(med_age_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))

## categorize pop12_sqmi with low density (rural) as high priority
sum_pop12_sqmi<-summary(demo_sub$pop12_sqmi)
#sum_pop12_sqmi
demo_sub<-demo_sub%>%
  mutate(pop12_sqmi_CAT= case_when(
    pop12_sqmi < as.numeric(pop12_sqmi[2]) ~ "High priority",      
    pop12_sqmi < as.numeric(pop12_sqmi[5]) ~ "Medium priority",
    TRUE ~ "Low priority"))%>%
  mutate(pop12_sqmi_CAT = factor(pop12_sqmi_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))

## create % renter (of all households) variable
demo_sub<-demo_sub%>%
  mutate(renter_ratio=renter_occ/(owner_occ+renter_occ))

## categorize renter_ratio with higher ratio being a higher priority
sum_renter_ratio<-summary(demo_sub$renter_ratio)
#sum_renter_ratio
demo_sub<-demo_sub%>%
  mutate(renter_ratio_CAT= case_when(
    renter_ratio < as.numeric(renter_ratio[2]) ~ "Low priority",      
    renter_ratio < as.numeric(renter_ratio[5]) ~ "Medium priority",
    TRUE ~ "High priority"))%>%
  mutate(renter_owner_ratio_CAT = factor(renter_ratio_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))

demo_col_order <- c("name", "pop2012","pop12_sqmi", "pop12_sqmi_CAT", "med_age", 
               "med_age_CAT", "renter_ratio", "renter_ratio_CAT")
demo_sub <- demo_sub[, demo_col_order]

```

```{r echo=FALSE, results=FALSE}
# BACKGROUND: mortality data
mort_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_county_mortality.csv",
                    stringsAsFactors = FALSE,na.strings = "")

## remove last 2 columns b/c mostly NAs ("Annotation_...")
mort_data<-mort_data[,1:8]

## column names are all capitalized
colnames(mort_data) <- str_to_lower(colnames(mort_data))

## filter by 'strata_name' as 'total_population'
mort_sub<-mort_data%>%
  filter(strata_name=="Total Population")
## Note: We kept occurrence and residence data because it's a metric of overall 
## population load at the hospital

## replace NAs with zeros
mort_sub <- mort_sub %>% mutate(count = ifelse(is.na(count), 0, count))

## filter non-chronic diseases
mort_sub<-mort_sub%>%
  filter(cause_desc %in% c("Alzheimer's disease", "Malignant neoplasms",
                     "Chronic lower respiratory diseases","Diabetes mellitus",
                     "Diseases of heart", 
                     "Essential hypertension and hypertensive renal disease",
                     "Chronic liver disease and cirrhosis", 
                     "Nephritis, nephrotic syndrome and nephrosis",
                     "Parkinson's disease", "Cerebrovascular diseases"))
## Note: We removed all non-chronic diseases: "all cause", "assault", 
## "accidents", "influenza" and "self-harm"

## summarize chronic death mortality by county
mort_sub_grouped<-mort_sub%>%
  group_by(county)%>%
  summarize(summed_chronic_dis_mort=sum(count))

```

```{r echo=FALSE, results=FALSE}
# BACKGROUND: healthcare (HCAI) data
healthcare_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/hcai_healthcare_construction.csv",
  stringsAsFactors = FALSE,na.strings = "")

#remove last column ("Collection.of.Counties") b/c mostly NAs
healthcare_data<-healthcare_data[,1:5]

#column names are all capitalized
colnames(healthcare_data) <- str_to_lower(colnames(healthcare_data))

#change "." to "_" for consistency with other data sets
names(healthcare_data) <- gsub(x = names(healthcare_data), pattern = "\\.", 
                               replacement = "_")  

#change county names to match other two data sets (remove numbers in front)
healthcare_data<-healthcare_data%>%
  mutate(county=substring(healthcare_data$county, 6))

## change money column ("Total.Costs...") to numeric
healthcare_data<-healthcare_data%>%
  mutate(total_costs_of_oshpd_projects = total_costs_of_oshpd_projects %>% 
           str_remove_all("[$,]"))
healthcare_data$total_costs_of_oshpd_projects<-
  as.numeric(healthcare_data$total_costs_of_oshpd_projects)

## fix "data_generation_date" (remove empty time)
healthcare_data$data_generation_date<-
  as.Date(healthcare_data$data_generation_date)

## filter data to only include projects with 'project status' as "Pending 
## Construction," "In Construction" or "In Closure"  
healthcare_sub<-healthcare_data%>%
  filter(oshpd_project_status!="In Review")
## Note: We only filtered out "in review" because all other projects with other
## statuses were guaranteed to be completed and we wanted to prioritize 
## locations with the fewest projects that were at any stage of completion

## filter out 2013 and 2014 data so we have 5 yrs of data (from 2015-2020)
healthcare_sub<-healthcare_sub%>%
  filter(data_generation_date > "2015-01-01")

## remove an unnecessary warning message: `summarise()` has grouped output by 
## 'county'. You can override using the `.groups` argument.
options(dplyr.summarise.inform = FALSE)

## average 'total_cost' in each category for each county to reduce 
## over-counting that is occurring in the data set
healthcare_sub_grouped<-healthcare_sub%>%
  group_by(county, oshpd_project_status)%>%
  summarize(summed_total_cost=mean(total_costs_of_oshpd_projects))

## R is forcing numbers to be in scientific notation, so fix
options(scipen = 999)

## summarize 'total_cost' of projects 'in closure', 'in construction', 
## or 'pending construction' over the 5 years (2015-2020)
healthcare_sub_grouped<-healthcare_sub%>%
  group_by(county)%>%
  summarize(summed_total_cost=sum(total_costs_of_oshpd_projects))

## make the total cost of all oshpd projects categorical where counties with  
## higher total costs are lower priority
s_summed_total_cost<-summary(healthcare_sub_grouped$summed_total_cost)
#s_summed_total_cost
healthcare_sub_grouped<-healthcare_sub_grouped%>%
  mutate(summed_total_cost_CAT= case_when(
    summed_total_cost < as.numeric(s_summed_total_cost[2]) ~ "High priority",      
    summed_total_cost < as.numeric(s_summed_total_cost[5]) ~ "Medium priority",
    TRUE ~ "Low priority"))%>%
  mutate(summed_total_cost_CAT = factor(summed_total_cost_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))

```


```{r echo=FALSE, results=FALSE}
# BACKGROUND: merge data sets 
merged_data<-full_join(mort_sub_grouped, healthcare_sub_grouped, by="county")
merged_data<-full_join(merged_data, demo_sub, by=c("county"= "name"))

## use merged data to relativize mortality data
merged_data<-merged_data%>%
  mutate(relative_chronic_dis_mort=(summed_chronic_dis_mort/pop2012))%>%
  select(-c(summed_chronic_dis_mort, pop2012))

## make relative_chronic_dis_mort categorical with higher counts of chronic 
## disease in hospitals being higher priority
sum_relative_chronic_dis_mort<-summary(merged_data$relative_chronic_dis_mort)
sum_relative_chronic_dis_mort
merged_data<-merged_data%>%
  mutate(relative_chronic_dis_mort_CAT= case_when(
    relative_chronic_dis_mort < 
      as.numeric(sum_relative_chronic_dis_mort[2]) ~ "Low priority",      
    relative_chronic_dis_mort < 
      as.numeric(sum_relative_chronic_dis_mort[5]) ~ "Medium priority",
    TRUE ~ "High priority"))%>%
  mutate(relative_chronic_dis_mort_CAT = factor(relative_chronic_dis_mort_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))


```


# **Problem Statement**
Rural hospitals are struggling to stay open throughout the country and the state
of California ranks #4 in rural hospital closures ([citation](https://kesq.com/stacker-california/2021/12/09/california-is-the-4-state-with-the-most-rural-hospital-closures-since-2005/)). When rural hospitals close, there is a
subsequent rise in mortality ([citation](https://www.nber.org/papers/w26182?utm_campaign=ntwh&utm_medium=email&utm_source=ntwg12)).
This is why the California Department of Public Health, Office of Health Equity
is excited that a new policy has just been created to fund a 
public-private partnership for healthcare facility improvement in 5 of 
California's rural counties that have received minimal funding from the 
Department of Health Care Access and Information (HCAI) over the past 5 years. 
Here we have analyzed 5 variables in order to determine which 5 counties are the
best targets for the development fund proposals.


# **Methods**
### **Data Sources**
To make recommendations on target counties, we used variables from three 
open-access, State of California data sources which are described below.

#### *[California demographics by County](https://github.com/PHW290/phw251_projectdata/blob/main/ca_county_demographic.csv)*
Includes data on population characteristics by county including average 
population density, counts by race/ethnicity, sex, median age, and housing data.

**Years and/or dates of data**

Data was collected in 2012.
 
**Description of cleaning and creating new variables**

To prepare the data, we:

1. Data cleaning
  + make "County" the first (left-most) column in the data set for merging
2. Data filtering
  + we subset data to only include needed data: "pop2012","pop12_sqmi", "name", "med_age","renter_occ", "owner_occ"
3. Data summarizing
  + *NEW VARIABLE*: categorize Median Age into "High Priority" (counties with Median Age above the third quantile), "Medium Priority" (counties with Median Age between the first and third quantiles), and "Low Priority" (counties with Median Age below the first quantile)
  + *NEW VARIABLE*: categorize Population per Square Mile (hereafter Population Density) into "High Priority" (counties with Population Density above the third quantile), "Medium Priority" (counties with Population Density between the first and third quantiles), and "Low Priority" (counties with Population Density below the first quantile)
  + *NEW VARIABLE*: create ratio of renting households as a fraction of total households.
  + *NEW VARIABLE*: categorize Percent of Household that were Renters (hereafter Percent Renters) into "High Priority" (counties with Percent Renters above the third quantile), "Medium Priority" (counties with Percent Renters between the first and third quantiles), and "Low Priority" (counties with Percent Renters below the first quantile)



#### *[California Department of Public Health Care: Death Profiles by County](https://data.ca.gov/dataset/death-profiles-by-county)*
Includes data on mortality events stratified by age, sex, race, hospital 
department, and residence status of the deceases relative to the hospital where
the mortality event occurred.

**Years and/or dates of data**

The data set contains data from 2014 through 2020, though the data
was restricted to data from 2015 through 2020 to correspond with requirement 
that the data used in this analysis was collected over the most recent 5 years.
 
**Description of cleaning and creating new variables**

To prepare the data, we:

1. Data cleaning
  + made all column names lowercase
  + replaced NAs with zeros, as instructed
2. Data filtering
  + we filtered data in "strata_name" to only include data about the Total Population (we removed age, sex, race and other stratifiers)
  + we kept both occurrence and residence data because together they are a metric of overall population load at the hospital
  + we filtered out non-chronic diseases (all cause, assault, accidents, influenza and self-harm)
  + we kept: Alzheimer's disease, Malignant neoplasms, Chronic lower respiratory diseases, Diabetes mellitus, Diseases of heart, Essential hypertension and hypertensive renal disease, Chronic liver disease and cirrhosis, Nephritis, nephrotic syndrome and nephrosis, Parkinson's disease, and Cerebrovascular diseases
  + removed data from 2013 so we have 5 yrs of available data (2015-2020) 
3. Data summarizing
  + *NEW VARIABLE*: sum all chronic disease mortality events by county
  + *NEW VARIABLE*: [after merging data sets] divide summed chronic disease mortality by average population size (from the demographic data set) for each county
  + *NEW VARIABLE*: [after merging data sets] categorize Relative Numbers of Chronic Disease Mortality (hereafter Mortality) into "High Priority" (counties with Mortality above the third quantile), "Medium Priority" (counties with Mortality between the first and third quantiles), and "Low Priority" (counties with Mortality below the first quantile)



#### *[Department of Health Care Access and Information: Total Construction Cost of Healthcare Projects](https://data.ca.gov/dataset/total-construction-cost-of-healthcare-projects)*
Includes number of projects and total spending on hospital projects funded by 
Department of Health Care Access and Information (HCAI) by county, date, and
project status ("in review", "pending construction", "in construction" and "in
closure.")

**Years and/or dates of data**

The data set contains data from 10-14-2013 through 08-11-2022, though the data
was restricted to 5 years from 2015 through 2020 to correspond with data 
availability in the other data sets.
 
**Description of cleaning and creating new variables**

To prepare the data, we:

1. Data cleaning
  + made all column names lowercase
  + replaced spaces with underscores
  + removed the numbers in front of the County names so they would match the format of the other data sets
  + removed dollar signs from Total Costs 
  + changed dates to a date format recognizable by R
2. Data filtering
  + removed projects listed as "in review" because we wanted to prioritize locations with the fewest projects that were at any stage of completion and all other projects with other statuses were guaranteed to be completed
  + removed data from 2013 and 2014 so we have 5 yrs of data - from 2015-2020 - to match other available data
3. Data summarizing
  + average Total Costs in each category for each county to reduce over-counting that is occurring in the data set
  + *NEW VARIABLE*: sum the Total Costs over the 4 categories ('in closure', 'in construction', or 'pending construction') for each county
  + *NEW VARIABLE*: categorize Total Costs into "High Priority" (counties with Total Costs below the first quantile), "Medium Priority" (counties with Total Costs between the first and third quantiles), and "Low Priority" (counties with Total Costs above the third quantile)

### **Analytic methods**
We decided to approach the ranking of counties with two different techniques to 
examine the robustness of our final recommendations. Our first approach was to 
use to two graphs. In the first graph would be a scatterplot where we would use 
just the 3 demographic variables to identify top priority counties with fraction 
of renters on the x-axis, median age of the residents on the y-axis and dots to 
color code populations as being high or low priority based on population density
(lower density was higher priority).

To prepare for our second scatterplot, we collapsed the ranking data from the 
first scatterplot into a single data set by coding a *NEW VARIABLE* where we 
gave 2 points to variables identified as "high priority," 1 point for variables 
identified as "medium priority" and zero points for "low priority" variables for 
each county. For example, Amador County is a high priority population based on
median age (2 points) and population density (2 more points) but a low priority 
population based on proportion of residents that were renters (0 points) for a 
total demographic ranking score of 4 points. This 6th variable was used to color 
code the dots in the second scatterplot where we had Total Cost of Previous and 
Ongoing OSHPD/HCAI project on the x-axis and Relative Levels of Chronic Disease 
Mortality Events on the y-axis. This second graph highlights the counties that 
we would consider recommending for future priority funding because it includes 
all 5 variables of interest.

Our second approach was to use a table with a *NEW VARIABLE* to rank the 
counties on all 5 variables of interest simultaneously. The ranking variable 
uses the same method that was previously explained but for all 5 variables 
instead of just the 3 demographic variables: 2 points were given to variables 
identified as "high priority," 1 point for variables identified as "medium 
priority" and zero points for "low priority" variables for each county. For 
example, Siskiyou County is a high priority population based on median age 
(2 points), population density (2 more points) and relative chronic disease 
mortality (additional 2 points). It was a medium priority population based on 
total previous OSHPD/HCAI funding (1 point) and a low priority population based 
on proportion of residents that were renters (0 points) for a total ranking 
score of 7 points. Counties were then ranked on their total ranking score.


# **Results**


### *Figure 1:*
add caption
```{r echo=FALSE, results=TRUE}
## renter ratio median = 39%
## median age median = 37.05
## population density 1st quantile (low cutoff) = 25.887
## population density 3rd quantile (high cutoff) = 333.485
ggplot(data = merged_data, aes(x = renter_ratio, y = med_age)) + 
geom_point(data = merged_data, aes(x = renter_ratio, y = med_age, 
                                   color = pop12_sqmi_CAT)) +
geom_text_repel(aes(label=ifelse((med_age > 37 & renter_ratio > 0.39 
    & (pop12_sqmi_CAT=="High priority"| pop12_sqmi_CAT=="Medium priority")), 
    county, "")))+
  labs(title = "Priority counties identified based on demographic data only:",
subtitle = "counties with high median age (>37yo), high ratio of renters (>39%), 
and low or medium population density (<333 people/sqmi)", 
       x = "Fraction of renters", 
       y = "Median age of county residents", 
       color = 
         bquote(atop(Population~per~mile^{"2"}, "rural as high priority")))+
  theme(plot.title=element_text(hjust=0.5),
        plot.subtitle=element_text(hjust=0.5))
                          
```



\newpage
### *Figure 2:*
add caption
# note: this figure should highlight top 5 counties for recommended funding
```{r echo=FALSE, results=TRUE}
# make data set with continuous data and ranking factor for the demographic 
# data in the first figure
second_fig_data_temp<-merged_data%>%
  select(c("county", "pop12_sqmi_CAT", "med_age_CAT", "renter_ratio_CAT"))%>%
 rowwise() %>%
 mutate(number_highs= sum(c_across(2:4) == "High priority", na.rm = TRUE),
        number_mediums= sum(c_across(2:4) == "Medium priority", na.rm = TRUE),
        demo_rank=(number_highs*2)+number_mediums
        )%>%
  ungroup()%>%
  select(c("county", "demo_rank"))
  
second_fig_data_final<-full_join(second_fig_data_temp, merged_data, by="county")
## summary(second_fig_data_final$relative_chronic_dis_mort)

# make the figure
## relative chronic disease mortality median = 0.07213
## log(relative chronic disease mortality median) = log(0.07213) = -2.629285
## summed total cost median = 5961782208
## log(summed total cost median) = log(5961782208) = 22.50864
ggplot(data = second_fig_data_final, 
       aes(y = relative_chronic_dis_mort, x = log(summed_total_cost))) + 
geom_point(data = second_fig_data_final, 
           aes(y = relative_chronic_dis_mort, x = log(summed_total_cost), 
                                   color = as.factor(demo_rank))) +
guides(color = guide_legend(reverse=TRUE))+
 geom_text_repel(aes(label=ifelse(
   (relative_chronic_dis_mort >= 0.07213 & summed_total_cost<=5961782208 
    & demo_rank >3), county, "")), max.overlaps = Inf)+
labs(title = "Priority counties identified with all data:",
 subtitle = "counties with high relative chronic disease mortality, 
low previous investment, and high priority based on demographics", 
        x = "Log(total dollars invested in previous projects)", 
        y = 
"Relative number of chronic disease \nmortality events between 2015-2020", 
        color = "Priority ranking \nscore based on \ndemographics data") +
   theme(plot.title=element_text(hjust=0.5),
         plot.subtitle=element_text(hjust=0.5))

```


\newpage
### *Table:* 
add caption
# note: table should highlight top 5 counties for recommended funding
Table is still being worked on as part of milestone 4.
```{r echo=FALSE, results=TRUE}
table_col_order <- c("county", "summed_total_cost", "pop12_sqmi",
                     "med_age", "renter_ratio",
                     "relative_chronic_dis_mort", "med_age_CAT",
                     "summed_total_cost_CAT", "pop12_sqmi_CAT",
                     "renter_ratio_CAT","relative_chronic_dis_mort_CAT")
merged_data_for_table <- merged_data[, table_col_order]

 table<-merged_data_for_table%>%
  rowwise() %>%
  mutate(number_highs= sum(c_across(7:11) == "High priority", na.rm = TRUE),
         number_mediums= sum(c_across(7:11) == "Medium priority", na.rm = TRUE),
         temp_rank=(number_highs*2)+number_mediums
         )%>%
   ungroup()%>%
   arrange(desc(temp_rank))%>%
   select(-c(number_highs, number_mediums))%>%
   slice(1:15)
table


# kable(table,
#        col.names = c("County","Chronic disease mortality burden",
#                      "Previous spending on projects",
#                      "Population density", "Median age of population",
#                      "% population that are renters"),
#        caption="Top 10 Counties ranked by need for oshpd projects.",
#        booktabs=TRUE,
#        align='lccccc')%>%
#   kable_styling(latex_options="scale_down")

```


# **Discussion**
(interpretation of results)
# note these interpretations are copy-pasted from milestone 4 - need to be edited
Plot 1: As a first step, we plotted demographic information only (population per 
square mile, renter to homeowner ratio, and median age of the residents of the 
county) and highlighted the counties that would be the highest priority if
we were only looking at the demographic data. We prioritized low density, higher
percentage of renters, and older populations.

Plot 2: Here we integrated demographic data from Plot 1 with additional data: 
the total dollars invested in prior projects (with higher priority for less 
money previously invested) and number of chronic disease mortality events 
between 2015-2020, relativized by total population (with higher priority being 
higher relative levels of chronic disease mortality). We used demographic data 
from Plot 1 to create a new ranked variable which we used to color code data in 
Plot 2. Counties highlighted in Plot 1 were ranked more highly in Plot 2. Thus 
this Plot includes all 5 variables of interest to identify the counties that 
require greater funding. Note that money invested in previous projects has been
logged to make the plot easier to read.

Table 1: We used a table as a different way to organize the 5 variables of 
interest. We categorized each variable and then ranked each as high, medium, or
low priority. We then created a new variable where "high priority" variables
were given two points, "medium priority" variables given 1 point, and "low
priority" variables given 0 points for each county. Then Counties are ranked
by number of points. Most of the Counties highlighted in Plot/Visualization 2 
were also the most highly ranked in the table.
