---
title: "R Project Milestone 2"
author: "Moyra Rasheed, Courtney Coon, Jarett Maycott"
date: '2022-09-30'
output:
  pdf_document: default

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readr)
library(dplyr)
library(stringr)
 
```


# Importing data
## demographic dataset
```{r}
demo_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_county_demographic.csv",
                    stringsAsFactors = FALSE)
str(demo_data)
#data is clean and ready to use
```

\newpage

## mortality dataset
```{r}
mort_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_county_mortality.csv",
                    stringsAsFactors = FALSE,na.strings = "")

#remove last 2 columns b/c mostly NAs ("Annotation_...")
mort_data<-mort_data[,1:8]
head(mort_data)

#column names are all capitalized
colnames(mort_data) <- str_to_lower(colnames(mort_data))
head(mort_data)
```

\newpage
## mortality dataset con't
```{r}
#interested in knowing categories of mortality
unique(mort_data$cause_desc)
unique(mort_data$year)

#note "strata" and "strata_name" may have categories we want to pull out at some point
```
\newpage


## HCAI Healthcare dataset
```{r}
healthcare_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/hcai_healthcare_construction.csv",
  stringsAsFactors = FALSE,na.strings = "")

#remove last column ("Collection.of.Counties") b/c mostly NAs
healthcare_data<-healthcare_data[,1:5]
head(healthcare_data)

#column names are all capitalized
colnames(healthcare_data) <- str_to_lower(colnames(healthcare_data))
head(healthcare_data)

```


\newpage
## HCAI Healthcare dataset con't
```{r}
#change "." to "_" for consistency with other datasets
names(healthcare_data) <- gsub(x = names(healthcare_data), pattern = "\\.", 
                               replacement = "_")  
head(healthcare_data)

#change county names to match other two datasets (remove numbers in front)
healthcare_data<-healthcare_data%>%
  mutate(county=substring(healthcare_data$county, 6))
head(healthcare_data)
```
\newpage
## HCAI Healthcare dataset con't
```{r}
#change money column ("Total.Costs...") to numeric
healthcare_data<-healthcare_data%>%
  mutate(total_costs_of_oshpd_projects = total_costs_of_oshpd_projects %>% 
           str_remove_all("[$,]"))
healthcare_data$total_costs_of_oshpd_projects<-as.numeric(healthcare_data$total_costs_of_oshpd_projects)
head(healthcare_data)

#fix "data_generation_date" (remove empty time)
healthcare_data$data_generation_date<-as.Date(healthcare_data$data_generation_date)
head(healthcare_data)

```

\newpage

# Description of dataset
All 3 data sets are CSV files. 


## What is the data source? How does the dataset relate to the group problem 
## statement and question?
### 1. Demographic: Census Data
Includes 58 observations with 22 variables. All are numeric or integers except
for the first column which is a character with county names.

### 2. Mortality: CA open data portal, California Department of Public Health 
Includes 147,784 observations with 10 variables. most are characters except for 
year and count of mortality.

### 3. Healthcare: CA open data portal, Department of Healthcare Access and Information
Includes 53,592 observations with 5 variables. 

#### How does the dataset relate to the group problem statement and question?
Allocated funding for healthcare  for vulnerable populations in California. 
Certain regions in California are considered to be vulnerable populations. 
Given that they need access to better funding, thereby providing access to 
better healthcare, we are able to allocate funds to certain regions. By 
utilizing all 3 data sets, we can focus on rural populations, seniors and 
renters, higher mortality rate and the counties that have average or below 
average funding available in the last 5 years.  

\newpage
# Identify data types for 5+ data elements/columns/variables
## Identify 5+ data elements required for your specified scenario. 
## If <5 elements are required to complete the analysis, please choose 
## additional variables of interest in the data set to explore in this milestone.
1. Demographic Data: demo_data$pop12_sqmi: Type:chr 
2. Demographic Data: demo_data$name: Type:chr
3. Demographic: demo_data$med_age: Type:num
4. Demographics: demo_data$renter_occ: Type:num
5. Demographics: demo_data$owner_occ: Type:num
6. Mortality: mort_data$county: Type:chr
7. Mortality: mort_data$year: Type:chr
8. Mortality: mort_data$cause_desc: Type: chr
9. Mortality: mort_data$count: Type:num
10. HCAI healthcare: healthcare_data$oshph_project_status: Type:chr
11. HCAI healthcare: healthcare_data$data_generation_date: Type:chr
12. HCAI healthcare:healthcare_data$total_costs_of_oshpd_projects: Type:num
13. HCAI healthcare: healthcare_data$county: Type:chr
14. HCAI healthcare: healthcare_data$number_of_oshpd_projects: Type:num


## Identify the desired type/format for each variable—will you need to convert 
## any columns to numeric or another type? 
### 1. HCAI: data_generation_date: change format from date-time to strictly date;
### may need to change to year to merge data for Milestone #3
### 2. HCAI: changed "total_cost_projects" from chr (string) to numeric 
### by removing "$"
### The remaining columns are in the appropiate format for now.

\newpage
# Provide a basic description of the 5+ data elements (Numeric: mean, median, 
# range; Character: unique values/categories)
```{r}
#1: demo_data$pop12_sqmi
summary(demo_data$pop12_sqmi)
```

```{r}
#2: demo_data$name
unique(demo_data$name)
```

```{r}
#3: demo_data$name
summary(demo_data$name)
```

```{r}
#4: demo_data$renter_occ
summary(demo_data$renter_occ)
```

```{r}
#5: demo_data$owner_occ
summary(demo_data$owner_occ)
```

\newpage
```{r}
#6: mort_data$county
unique(mort_data$county)
```

```{r}
#7: mort_data$year
unique(mort_data$year)
```

```{r}
#8: mort_data$cause_desc
unique(mort_data$cause_desc)
```

```{r}
#9: mort_data$count
summary(mort_data$count)
```

```{r}
#10: healthcare_data$oshpd_project_status
unique(healthcare_data$oshpd_project_status)
```

```{r}
#11: healthcare_data$data_generation_date
unique(healthcare_data$data_generation_date)
```

```{r}
#12: healthcare_data$total_costs_of_oshpd_projects
summary(healthcare_data$total_costs_of_oshpd_projects)
```

```{r}
#13: healthcare_data$county
unique(healthcare_data$county)
```

```{r}
#14: healthcare_data$number_of_oshpd_projects
summary(healthcare_data$number_of_oshpd_projects)
```
