---
title: "R Project Milestone 3"
author: "Moyra Rasheed, Courtney Coon, Jarett Maycott"
date: '2022-11-26'
output:
  pdf_document: default

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readr)
library(dplyr)
library(stringr)
 
```

```{r echo=FALSE, results=FALSE}
# Prep demographic data
## read in data
demo_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_county_demographic.csv",
                    stringsAsFactors = FALSE)

## subset demographic dataset to include only needed columns
demo_sub<-select(demo_data, c("pop2012","pop12_sqmi", "name", "med_age","renter_occ", "owner_occ"))

## categorize median age: younger is lower priority, older is high priority
sum_med_age<-summary(demo_sub$med_age)
sum_med_age
demo_sub<-demo_sub%>%
  mutate(med_age_CAT= case_when(
    med_age < as.numeric(sum_med_age[2]) ~ "Low priority",      
    med_age < as.numeric(sum_med_age[5]) ~ "Medium priority",
    TRUE ~ "High priority"))%>% 
  mutate(med_age_CAT = factor(med_age_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))

## categorize pop12_sqmi with low density (rural) as high priority
sum_pop12_sqmi<-summary(demo_sub$pop12_sqmi)
sum_pop12_sqmi
demo_sub<-demo_sub%>%
  mutate(pop12_sqmi_CAT= case_when(
    pop12_sqmi < as.numeric(pop12_sqmi[2]) ~ "High priority",      
    pop12_sqmi < as.numeric(pop12_sqmi[5]) ~ "Medium priority",
    TRUE ~ "Low priority"))%>%
  mutate(pop12_sqmi_CAT = factor(pop12_sqmi_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))

## create % renter (of all households) variable
demo_sub<-demo_sub%>%
  mutate(renter_ratio=renter_occ/(owner_occ+renter_occ))

## categorize renter_ratio with higher ratio being a higher priority
sum_renter_ratio<-summary(demo_sub$renter_ratio)
sum_renter_ratio
demo_sub<-demo_sub%>%
  mutate(renter_ratio_CAT= case_when(
    renter_ratio < as.numeric(renter_ratio[2]) ~ "Low priority",      
    renter_ratio < as.numeric(renter_ratio[5]) ~ "Medium priority",
    TRUE ~ "High priority"))%>%
  mutate(renter_owner_ratio_CAT = factor(renter_ratio_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))

demo_col_order <- c("name", "pop2012","pop12_sqmi", "pop12_sqmi_CAT", "med_age", 
               "med_age_CAT", "renter_ratio", "renter_ratio_CAT")
demo_sub <- demo_sub[, demo_col_order]

```

```{r echo=FALSE, results=FALSE}
# Prep mortality data
mort_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_county_mortality.csv",
                    stringsAsFactors = FALSE,na.strings = "")

## remove last 2 columns b/c mostly NAs ("Annotation_...")
mort_data<-mort_data[,1:8]

## column names are all capitalized
colnames(mort_data) <- str_to_lower(colnames(mort_data))

## filter by 'strata_name' as 'total_population'
mort_sub<-mort_data%>%
  filter(strata_name=="Total Population")
## Note: We kept occurrence and residence data because it's a metric of overall 
## population load at the hospital

## replace NAs with zeros
mort_sub <- mort_sub %>% mutate(count = ifelse(is.na(count), 0, count))

## filter non-chronic diseases
mort_sub<-mort_sub%>%
  filter(cause_desc %in% c("Alzheimer's disease", "Malignant neoplasms",
                     "Chronic lower respiratory diseases","Diabetes mellitus",
                     "Diseases of heart", 
                     "Essential hypertension and hypertensive renal disease",
                     "Chronic liver disease and cirrhosis", 
                     "Nephritis, nephrotic syndrome and nephrosis",
                     "Parkinson's disease", "Cerebrovascular diseases"))
## Note: We removed all non-chronic diseases: "all cause", "assault", 
## "accidents", "influenza" and "self-harm"

## summarize chronic death mortality by county
mort_sub_grouped<-mort_sub%>%
  group_by(county)%>%
  summarize(summed_chronic_dis_mort=sum(count))

```

```{r echo=FALSE, results=FALSE}
# Prep the healthcare (HCAI) data
healthcare_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/hcai_healthcare_construction.csv",
  stringsAsFactors = FALSE,na.strings = "")

#remove last column ("Collection.of.Counties") b/c mostly NAs
healthcare_data<-healthcare_data[,1:5]

#column names are all capitalized
colnames(healthcare_data) <- str_to_lower(colnames(healthcare_data))

#change "." to "_" for consistency with other data sets
names(healthcare_data) <- gsub(x = names(healthcare_data), pattern = "\\.", 
                               replacement = "_")  

#change county names to match other two data sets (remove numbers in front)
healthcare_data<-healthcare_data%>%
  mutate(county=substring(healthcare_data$county, 6))

## change money column ("Total.Costs...") to numeric
healthcare_data<-healthcare_data%>%
  mutate(total_costs_of_oshpd_projects = total_costs_of_oshpd_projects %>% 
           str_remove_all("[$,]"))
healthcare_data$total_costs_of_oshpd_projects<-
  as.numeric(healthcare_data$total_costs_of_oshpd_projects)

## fix "data_generation_date" (remove empty time)
healthcare_data$data_generation_date<-
  as.Date(healthcare_data$data_generation_date)

## filter data to only include projects with 'project status' as "Pending 
## Construction," "In Construction" or "In Closure"  
healthcare_sub<-healthcare_data%>%
  filter(oshpd_project_status!="In Review")
## Note: We only filtered out "in review" because all other projects with other
## statuses were guaranteed to be completed and we wanted to prioritize 
## locations with the fewest projects that were at any stage of completion

## filter out 2013 and 2014 data so we have 5 yrs of data (from 2015-2020)
healthcare_sub<-healthcare_sub%>%
  filter(data_generation_date > "2015-01-01")

## remove an unnecessary warning message: `summarise()` has grouped output by 
## 'county'. You can override using the `.groups` argument.
options(dplyr.summarise.inform = FALSE)

## average 'total_cost' in each category for each county to reduce 
## over-counting that is occurring in the data set
healthcare_sub_grouped<-healthcare_sub%>%
  group_by(county, oshpd_project_status)%>%
  summarize(summed_total_cost=mean(total_costs_of_oshpd_projects))

## R is forcing numbers to be in scientific notation, so fix
options(scipen = 999)

## summarize 'total_cost' of projects 'in closure', 'in construction', 
## or 'pending construction' over the 5 years (2015-2020)
healthcare_sub_grouped<-healthcare_sub%>%
  group_by(county)%>%
  summarize(summed_total_cost=sum(total_costs_of_oshpd_projects))

## make the total cost of all oshpd projects categorical where counties with  
## higher total costs are lower priority
s_summed_total_cost<-summary(healthcare_sub_grouped$summed_total_cost)
s_summed_total_cost
healthcare_sub_grouped<-healthcare_sub_grouped%>%
  mutate(summed_total_cost_CAT= case_when(
    summed_total_cost < as.numeric(s_summed_total_cost[2]) ~ "High priority",      
    summed_total_cost < as.numeric(s_summed_total_cost[5]) ~ "Medium priority",
    TRUE ~ "Low priority"))%>%
  mutate(summed_total_cost_CAT = factor(summed_total_cost_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))

```


```{r echo=FALSE, results=FALSE}
#merge data sets by county
merged_data<-full_join(mort_sub_grouped, healthcare_sub_grouped, by="county")
merged_data<-full_join(merged_data, demo_sub, by=c("county"= "name"))

## use merged data to relativize mortality data
merged_data<-merged_data%>%
  mutate(relative_chronic_dis_mort=(summed_chronic_dis_mort/pop2012))%>%
  select(-c(summed_chronic_dis_mort, pop2012))

## make relative_chronic_dis_mort categorical with higher counts of chronic 
## disease in hospitals being higher priority
sum_relative_chronic_dis_mort<-summary(merged_data$relative_chronic_dis_mort)
sum_relative_chronic_dis_mort
merged_data<-merged_data%>%
  mutate(relative_chronic_dis_mort_CAT= case_when(
    relative_chronic_dis_mort < 
      as.numeric(sum_relative_chronic_dis_mort[2]) ~ "Low priority",      
    relative_chronic_dis_mort < 
      as.numeric(sum_relative_chronic_dis_mort[5]) ~ "Medium priority",
    TRUE ~ "High priority"))%>%
  mutate(relative_chronic_dis_mort_CAT = factor(relative_chronic_dis_mort_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))


```



```{r echo=FALSE, results=FALSE}
library(ggplot2)
library(ggrepel)
```

# Visualization 1:
Plot 1: As a first step, we plotted demographic information only (population per 
square mile, renter to homeowner ratio, and median age of the residents of the 
county) and highlighted the counties that would be the highest priority if
we were only looking at the demographic data. We prioritized low density, higher
percentage of renters, and older populations.

```{r}
## renter ratio median = 39%
## median age median = 37.05
## population density 1st quantile (low cutoff) = 25.887
## population density 3rd quantile (high cutoff) = 333.485
ggplot(data = merged_data, aes(x = renter_ratio, y = med_age)) + 
geom_point(data = merged_data, aes(x = renter_ratio, y = med_age, 
                                   color = pop12_sqmi_CAT)) +
geom_text_repel(aes(label=ifelse((med_age > 37 & renter_ratio > 0.39 
    & (pop12_sqmi_CAT=="High priority"| pop12_sqmi_CAT=="Medium priority")), 
    county, "")))+
  labs(title = "Priority counties identified based on demographic data only:",
subtitle = "counties with high median age (>37yo), high ratio of renters (>39%), 
and low or medium population density (<333 people/sqmi)", 
       x = "Ratio of renters to homeowners", 
       y = "Median age of county residents", 
       color = 
         bquote(atop(Population~per~mile^{"2"}, "rural as high priority")))+
  theme(plot.title=element_text(hjust=0.5),
        plot.subtitle=element_text(hjust=0.5))
                          
```



\newpage
# Visualization 2:
Plot 2: Here we integrated demographic data from Plot 1 with additional data: 
the total dollars invested in prior projects (with higher priority for less 
money previously invested) and number of chronic disease mortality events 
between 2015-2020, relativized by total population (with higher priority being 
higher relative levels of chronic disease mortality). We used demographic data 
from Plot 1 to create a new ranked variable which we used to color code data in 
Plot 2. Counties highlighted in Plot 1 were ranked more highly in Plot 2. Thus 
this Plot includes all 5 variables of interest to identify the counties that 
require greater funding. Note that money invested in previous projects has been
logged to make the plot easier to read.

```{r}
# make data set with continuous data and ranking factor for the demographic 
# data in the first figure
second_fig_data_temp<-merged_data%>%
  select(c("county", "pop12_sqmi_CAT", "med_age_CAT", "renter_ratio_CAT"))%>%
 rowwise() %>%
 mutate(number_highs= sum(c_across(2:4) == "High priority", na.rm = TRUE),
        number_mediums= sum(c_across(2:4) == "Medium priority", na.rm = TRUE),
        demo_rank=(number_highs*2)+number_mediums
        )%>%
  ungroup()%>%
  select(c("county", "demo_rank"))
  
second_fig_data_final<-full_join(second_fig_data_temp, merged_data, by="county")
## summary(second_fig_data_final$relative_chronic_dis_mort)

# make the figure
## relative chronic disease mortality median = 0.07213
## log(relative chronic disease mortality median) = log(0.07213) = -2.629285
## summed total cost median = 5961782208
## log(summed total cost median) = log(5961782208) = 22.50864
ggplot(data = second_fig_data_final, 
       aes(y = relative_chronic_dis_mort, x = log(summed_total_cost))) + 
geom_point(data = second_fig_data_final, 
           aes(y = relative_chronic_dis_mort, x = log(summed_total_cost), 
                                   color = as.factor(demo_rank))) +
guides(color = guide_legend(reverse=TRUE))+
 geom_text_repel(aes(label=ifelse(
   (relative_chronic_dis_mort >= 0.07213 & summed_total_cost<=5961782208 
    & demo_rank >3), county, "")), max.overlaps = Inf)+
labs(title = "Priority counties identified with all data:",
 subtitle = "counties with high relative chronic disease mortality, 
low previous investment, and high priority based on demographics", 
        x = "Log(total dollars invested in previous projects)", 
        y = 
"Relative number of chronic disease \n mortality events between 2015-2020", 
        color = "Priority ranking \nscore based on \ndemographics data") +
   theme(plot.title=element_text(hjust=0.5),
         plot.subtitle=element_text(hjust=0.5))

```


\newpage
# Visualization 3: 
Table 1: We used a table as a different way to organize the 5 variables of 
interest. We categorized each variable and then ranked each as high, medium, or
low priority. We then created a new variable where "high priority" variables
were given two points, "medium priority" variables given 1 point, and "low
priority" variables given 0 points for each county. Then Counties are ranked
by number of points. Most of the Counties highlighted in Plot/Visualization 2 
were also the most highly ranked in the table.
```{r}
library(kableExtra)

table_col_order <- c("county", "summed_total_cost", "pop12_sqmi",
                     "med_age", "renter_ratio",
                     "relative_chronic_dis_mort", "med_age_CAT",
                     "summed_total_cost_CAT", "pop12_sqmi_CAT",
                     "renter_ratio_CAT","relative_chronic_dis_mort_CAT")
merged_data_for_table <- merged_data[, table_col_order]

 table<-merged_data_for_table%>%
  rowwise() %>%
  mutate(number_highs= sum(c_across(7:11) == "High priority", na.rm = TRUE),
         number_mediums= sum(c_across(7:11) == "Medium priority", na.rm = TRUE),
         temp_rank=(number_highs*2)+number_mediums
         )%>%
   ungroup()%>%
   arrange(desc(temp_rank))%>%
   select(-c(number_highs, number_mediums))%>%
   slice(1:15)
table

table <- table %>% select(, c(1,2,8,3,9,4,7,5,10,6,11))

table_printed = kable(table,
        col.names = c("","","",
                      "", "",
                      "", "",
                       "", "",
                      "", ""
                    ),
       caption="Top 10 Counties ranked by need for oshpd projects.",
        booktabs=TRUE,
        align='lclclclclcl')%>%
   kable_styling(latex_options="scale_down")

add_header_above(table_printed, c("County" = 1,"Previous spending on projects" = 2,
                      "Population density" = 2,
                      "Median age of population" = 2,
                       "% population that are renters"= 2,
                      "Chronic disease mortality burden" = 2
                    ))

```


