---
title: "R Project Milestone 2"
author: "Moyra Rasheed, Courtney Coon, Jarett Maycott"
date: '2022-09-30'
output:
  pdf_document: default

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readr)
library(dplyr)
library(stringr)
 
```


# Importing data
## demographic dataset
```{r}
demo_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_county_demographic.csv",
                    stringsAsFactors = FALSE)
str(demo_data)
#data is clean and ready to use
```

# The question I have(Moyra  : are we supposed to drop the rest of the columns that are not required? or do we keep all of them ? 
#The ones that we need" low population per square mile "pop12-sqm1'
#Med-age
#renters and homeowners (A high proportion of renters versus homeowners - we need to make a third variable )


\newpage

## mortality dataset
```{r}
mort_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_county_mortality.csv",
                    stringsAsFactors = FALSE,na.strings = "")

#remove last 2 columns b/c mostly NAs ("Annotation_...")
mort_data<-mort_data[,1:8]
head(mort_data)

#column names are all capitalized
colnames(mort_data) <- str_to_lower(colnames(mort_data))
head(mort_data)

#interested in knowing categories of mortality
unique(mort_data$cause_desc)
unique(mort_data$year)

#note "strata" and "strata_name" may have categories we want to pull out at some point
```
\newpage


## HCAI Healthcare dataset
```{r}
healthcare_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/hcai_healthcare_construction.csv",
  stringsAsFactors = FALSE,na.strings = "")

#remove last column ("Collection.of.Counties") b/c mostly NAs
healthcare_data<-healthcare_data[,1:5]
head(healthcare_data)

#column names are all capitalized
colnames(healthcare_data) <- str_to_lower(colnames(healthcare_data))
head(healthcare_data)

```


\newpage
## HCAI Healthcare dataset con't
```{r}
#change "." to "_" for consistency with other datasets
names(healthcare_data) <- gsub(x = names(healthcare_data), pattern = "\\.", 
                               replacement = "_")  
head(healthcare_data)

#change county names to match other two datasets (remove numbers in front)
healthcare_data<-healthcare_data%>%
  mutate(county=substring(healthcare_data$county, 6))
head(healthcare_data)
```
\newpage
## HCAI Healthcare dataset con't
```{r}
#change money column ("Total.Costs...") to numeric
healthcare_data<-healthcare_data%>%
  mutate(total_costs_of_oshpd_projects = total_costs_of_oshpd_projects %>% 
           str_remove_all("[$,]"))
healthcare_data$total_costs_of_oshpd_projects<-as.numeric(healthcare_data$total_costs_of_oshpd_projects)
head(healthcare_data)

#fix "data_generation_date" (remove empty time)
healthcare_data$data_generation_date<-as.Date(healthcare_data$data_generation_date)
head(healthcare_data)

```
#Jarrett/Moyra: do we just need closures for oshpd project status ? 

\newpage

# Description of dataset
All 3 data sets are CSV files. 


## What is the data source? How does the dataset relate to the group problem 
## statement and question?
### 1. Demographic: Census Data
Includes 58 observations with 22 variables. All are numeric or integers except
for the first column which is a character with county names.

### 2. Mortality: CA open data portal, California Department of Public Health 
Includes 147,784 observations with 10 variables. most are characters except for 
year and count of mortality.

### 3. Healthcare: CA open data portal, Department of Healthcare Access and Information
Includes 53,592 observations with 5 variables. 

#### How does the dataset relate to the group problem statement and question?
Allocated funding for healthcare  for vulnerable populations in California . 
Certain regions in California are considered to be vulnerable populations. Given that they need access to better funding, thereby providing access to better healthcare , we are able to allocate funds to certain regions . By utilizing all 3 data sets , we can focus on rural populations , seniors and renters , higher mortality rate and the counties that have average or below average funding available in the last 5 years.  

# Identify data types for 5+ data elements/columns/variables
#Identify 5+ data elements required for your specified scenario. If <5 elements are required to complete the analysis, please choose additional variables of interest in the data set to explore in this milestone.
### 1. Demographic Data : Population /:Type-Chr 
### 2. Demographic Data :name: /Type: Chr
### 3:Demographic: median_age/Type: Num
### 4.Demographics: Renter/Type: Num
### 5.Dermographics: Homeowner/Type:Num
### 6. Mortality: county: chr
### 7. Mortality: year: Type: chr
### 8. Mortality: Cause Mortality :Type:  chr
### 9. Mortality: Count: Type: chr
### 10.HCAI healthcare: oshph_project_status: Type: chr
### 11.HCAI healthcare: Date:Type: chr
### 12.HCAI healthcare: Total Costs:Type:Num
### 13.HCAI healthcare:County:Type:chr
###14.HCAI healthcare:: No of Oshpd projects:Type: Num


  
\newpage

## Identify the desired type/format for each variable—will you need to convert 
## any columns to numeric or another type? 
### 1. Mortality Data set: dropped last 2 columns that comprised of mainly "NA's) 
### 2. HCAI: changed "total_cost_projects" from chr to numeric 
### 3. 
### 4. 
### 5.   
### 6.   
### 7.   

## Provide a basic description of the 5+ data elements (Numeric: mean, median, 
## range; Character: unique values/categories)
### 1. 
### 2. 
### 3. 
### 4. 
### 5. 
### 6.   
### 7.   


