---
title: "R Project Milestone 3"
author: "Moyra Rasheed, Courtney Coon, Jarett Maycott"
date: '2022-10-28'
output:
  pdf_document: default

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readr)
library(dplyr)
library(stringr)
 
```

```{r echo=FALSE, results=FALSE}
demo_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_county_demographic.csv",
                    stringsAsFactors = FALSE)
#str(demo_data)
#data is clean and ready to use
```

```{r echo=FALSE, results=FALSE}
mort_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_county_mortality.csv",
                    stringsAsFactors = FALSE,na.strings = "")

#remove last 2 columns b/c mostly NAs ("Annotation_...")
mort_data<-mort_data[,1:8]
head(mort_data)

#column names are all capitalized
colnames(mort_data) <- str_to_lower(colnames(mort_data))
head(mort_data)

#note "strata" and "strata_name" may have categories we want to pull out at some point
```

```{r echo=FALSE, results=FALSE}
healthcare_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/hcai_healthcare_construction.csv",
  stringsAsFactors = FALSE,na.strings = "")

#remove last column ("Collection.of.Counties") b/c mostly NAs
healthcare_data<-healthcare_data[,1:5]
head(healthcare_data)

#column names are all capitalized
colnames(healthcare_data) <- str_to_lower(colnames(healthcare_data))
head(healthcare_data)

#change "." to "_" for consistency with other datasets
names(healthcare_data) <- gsub(x = names(healthcare_data), pattern = "\\.", 
                               replacement = "_")  
head(healthcare_data)

#change county names to match other two datasets (remove numbers in front)
healthcare_data<-healthcare_data%>%
  mutate(county=substring(healthcare_data$county, 6))
head(healthcare_data)

#change money column ("Total.Costs...") to numeric
healthcare_data<-healthcare_data%>%
  mutate(total_costs_of_oshpd_projects = total_costs_of_oshpd_projects %>% 
           str_remove_all("[$,]"))
healthcare_data$total_costs_of_oshpd_projects<-as.numeric(healthcare_data$total_costs_of_oshpd_projects)
head(healthcare_data)

#fix "data_generation_date" (remove empty time)
healthcare_data$data_generation_date<-as.Date(healthcare_data$data_generation_date)
head(healthcare_data)

#change money column ("Total.Costs...") to numeric
healthcare_data<-healthcare_data%>%
  mutate(total_costs_of_oshpd_projects = total_costs_of_oshpd_projects %>% 
           str_remove_all("[$,]"))
healthcare_data$total_costs_of_oshpd_projects<-as.numeric(healthcare_data$total_costs_of_oshpd_projects)
head(healthcare_data)

#fix "data_generation_date" (remove empty time)
healthcare_data$data_generation_date<-as.Date(healthcare_data$data_generation_date)
head(healthcare_data)

```

# Demographic data
```{r}
#subset demographic dataset to include only needed columns
demo_sub<-select(demo_data, c("pop12_sqmi", "name", "med_age","renter_occ", "owner_occ"))

#categorize median age: younger is lower priority, older is high priority
sum_med_age<-summary(demo_sub$med_age)
sum_med_age
demo_sub<-demo_sub%>%
  mutate(med_age_CAT= case_when(
    med_age < as.numeric(sum_med_age[2]) ~ "Low priority",      
    med_age < as.numeric(sum_med_age[5]) ~ "Medium priority",
    TRUE ~ "High priority"))%>% 
  mutate(med_age_CAT = factor(med_age_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))

#categorize pop12_sqmi with low density (rural) as high priority
sum_pop12_sqmi<-summary(demo_sub$pop12_sqmi)
sum_pop12_sqmi
demo_sub<-demo_sub%>%
  mutate(pop12_sqmi_CAT= case_when(
    pop12_sqmi < as.numeric(pop12_sqmi[2]) ~ "High priority",      
    pop12_sqmi < as.numeric(pop12_sqmi[5]) ~ "Medium priority",
    TRUE ~ "Low priority"))%>%
  mutate(pop12_sqmi_CAT = factor(pop12_sqmi_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))

#create % renter (of all households) variable
demo_sub<-demo_sub%>%
  mutate(renter_ratio=renter_occ/(owner_occ+renter_occ))

#categorize renter_ratio with higher ratio being a higher priority
sum_renter_ratio<-summary(demo_sub$renter_ratio)
sum_renter_ratio
demo_sub<-demo_sub%>%
  mutate(renter_ratio_CAT= case_when(
    renter_ratio < as.numeric(renter_ratio[2]) ~ "Low priority",      
    renter_ratio < as.numeric(renter_ratio[5]) ~ "Medium priority",
    TRUE ~ "High priority"))%>%
  mutate(renter_owner_ratio_CAT = factor(renter_ratio_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))

demo_sub<-select(demo_sub, c("pop12_sqmi_CAT", "name", "med_age_CAT","renter_ratio_CAT"))

```

\newpage
# Mortality data
```{r}
#filter by 'strata_name' as "total_population'
mort_sub<-mort_data%>%
  filter(strata_name=="Total Population")
#Note: We kept occurence and residence data because it's a metric of overall 
#population load at the hospital

#replace NAs with zeros
mort_sub <- mort_sub %>% mutate(count = ifelse(is.na(count), 0, count))

#filter non-chronic diseases
unique(mort_data$cause_desc)
#Note: We removed all non-chronic diseases: "all cause", "assault", "accidents",
#"influenza" and "self-harm"

mort_sub<-mort_sub%>%
  filter(cause_desc %in% c("Alzheimer's disease", "Malignant neoplasms",
                     "Chronic lower respiratory diseases","Diabetes mellitus",
                     "Diseases of heart", "Essential hypertension and hypertensive renal disease",
                     "Chronic liver disease and cirrhosis", "Nephritis, nephrotic syndrome and nephrosis",
                     "Parkinson's disease", "Cerebrovascular diseases"))
unique(mort_sub$cause_desc)

#summarize chronic death mortality by county
mort_sub_grouped<-mort_sub%>%
  group_by(county)%>%
  summarize(summed_chronic_dis_mort=sum(count))

#make summed_chronic_dis_mort categorical with higher counts of chronic disease
#in hospitals being higher priority
sum_summed_chronic_dis_mort<-summary(mort_sub_grouped$summed_chronic_dis_mort)
sum_summed_chronic_dis_mort
mort_sub_grouped<-mort_sub_grouped%>%
  mutate(summed_chronic_dis_mort_CAT= case_when(
    summed_chronic_dis_mort < as.numeric(sum_summed_chronic_dis_mort[2]) ~ "Low priority",      
    summed_chronic_dis_mort < as.numeric(sum_summed_chronic_dis_mort[5]) ~ "Medium priority",
    TRUE ~ "High priority"))%>%
  mutate(summed_chronic_dis_mort_CAT = factor(summed_chronic_dis_mort_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))


```

\newpage
# Healthcare data
```{r}
#filter data to only include projects with 'project status' as "Pending 
#Construction," "In Construction" or "In Closure"  
healthcare_sub<-healthcare_data%>%
  filter(oshpd_project_status!="In Review")
#Note: We only filtered out "in review" because all other projects with other
#statuses were guaranteed to be completed and we wanted to prioritize locations
#with the fewest projects that were at any stage of completion

#filter out 2013 data
healthcare_sub<-healthcare_sub%>%
  filter(data_generation_date > "2015-01-01")

#summarize 'total_cost' of projects 'in closure' over the 5 years (2015-2020)
#FIX: PRIORITIZE LOW COST AND LOW NUMBER OF PROJECTS
healthcare_sub_grouped<-healthcare_sub%>%
  group_by(county)%>%
  summarize(summed_total_cost=sum(total_costs_of_oshpd_projects))

#make the total cost of all oshpd projects categorical where counties with higher 
#total costs are lower priority
s_summed_total_cost<-summary(healthcare_sub_grouped$summed_total_cost)
s_summed_total_cost
healthcare_sub_grouped<-healthcare_sub_grouped%>%
  mutate(summed_total_cost_CAT= case_when(
    summed_total_cost < as.numeric(s_summed_total_cost[2]) ~ "High priority",      
    summed_total_cost < as.numeric(s_summed_total_cost[5]) ~ "Medium priority",
    TRUE ~ "Low priority"))%>%
  mutate(summed_total_cost_CAT = factor(summed_total_cost_CAT, 
    levels = c("Low priority", "Medium priority", "High priority")))

```

\newpage
# Merge the 3 datasets
```{r}
#merge data sets by county
merged_data<-full_join(mort_sub_grouped, healthcare_sub_grouped, by="county")
merged_data<-full_join(merged_data, demo_sub, by=c("county"= "name"))
merged_data <- subset( merged_data, select = -c(summed_chronic_dis_mort, summed_total_cost))
head(merged_data)

```

\newpage
# Data Dictionary
## variable 1: "county"
The county the data comes from.

## variable 2: "summed_chronic_dis_mort_CAT"
This is the summed number of mortality cases from chronic diseases in each county 
from 2015-2020. It was categorized into low, medium and high priority based on 
whether density was below the 1st quantile (low priority), between the 1st and 
3rd quantile (medium priority), or above the 3rd quantile (high priority) because
we wanted to priority counties with higher burdens of chronic diseases.

## variable 3: "summed_total_cost_CAT"
This is the total cost of all projects "Pending Construction," "In Construction,"
or "In Closure" from 1/1/2015 through 8/11/2022 (most recent data). It was 
categorized into low, medium and high priority based on whether summed total costs 
were below the 1st quantile (high priority), between the 1st and 3rd quantile 
(medium priority), or above the 3rd quantile (low priority) because we wanted to
prioritize counties that had received less funding for oshpd projects.

## variable 4: "pop12_sqmi_CAT"
This is population density at the county level categorized into low, medium and 
high priority based on whether density was below the 1st quantile (high priority),
between the 1st and 3rd quantile (medium priority), or above the 3rd quantile 
(low priority) because we wanted to prioritize rural counties i.e., those with
lower population densities.

## variable 5: "med_age_CAT"
This is median age of residents at the county level categorized into low, medium 
and high priority based on whether median age was below the 1st quantile (low 
priority), between the 1st and 3rd quantile (medium priority), or above the 3rd 
quantile (high priority) because we wanted to prioritize counties with more
elderly populations.

## variable 6: "renter_ratio_CAT"
This variable was created by dividing the number of renters by the total number
households (which was equal to renter_occ and owner_occ) in a county and then 
categorizing them into low, medium and high priority. The ratio was categorized 
as high priority if it was below the 1st quantile, medium priority if it was 
between 1st and 3rd quantile, or low priority if it was above the 3rd quantile 
as compared to other counties in the state of California.

\newpage
# Table with descriptive stats for variables in data dictionary
```{r}
library(kableExtra)

table<-merged_data%>%
 rowwise() %>%
 mutate(number_highs= sum(c_across(2:6) == "High priority", na.rm = TRUE),
        number_mediums= sum(c_across(2:6) == "Medium priority", na.rm = TRUE),
        temp_rank=(number_highs*2)+number_mediums
        )%>%
  ungroup()%>%
  arrange(desc(temp_rank))%>%
  select(-c(number_highs, number_mediums,temp_rank))%>%
  slice(1:10)
head(table)

kable(table,
       col.names = c("County","Chronic disease mortality burden",
                     "Previous spending on projects", 
                     "Population density", "Median age of population", 
                     "% population that are renters"),
       caption="Top 10 Counties ranked by need for oshpd projects.",
       booktabs=TRUE, 
       align='lccccc')%>%
  kable_styling(latex_options="scale_down")

 table
```





