---
title: "R Project Milestone 2"
author: "Moyra Rasheed, Courtney Coon, Jarett Maycott"
date: '2022-09-30'
output:
  pdf_document: default

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readr)
library(dplyr)
library(stringr)
#ok 
```


# Importing data
## demographic dataset
```{r}
demo_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_county_demographic.csv",
                    stringsAsFactors = FALSE)
str(demo_data)
#data is clean and ready to use
```
\newpage

## mortality dataset
```{r}
mort_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/ca_county_mortality.csv",
                    stringsAsFactors = FALSE,na.strings = "")

#remove last 2 columns b/c mostly NAs ("Annotation_...")
mort_data<-mort_data[,1:8]
head(mort_data)

#column names are all capitalized
colnames(mort_data) <- str_to_lower(colnames(mort_data))
head(mort_data)

#interested in knowing categories of mortality
unique(mort_data$cause_desc)
unique(mort_data$year)

#note "strata" and "strata_name" may have categories we want to pull out at some point
```
\newpage


## HCAI Healthcare dataset
```{r}
healthcare_data<-read.csv(
  "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/hcai_healthcare_construction.csv",
  stringsAsFactors = FALSE,na.strings = "")

#remove last column ("Collection.of.Counties") b/c mostly NAs
healthcare_data<-healthcare_data[,1:5]
head(healthcare_data)

#column names are all capitalized
colnames(healthcare_data) <- str_to_lower(colnames(healthcare_data))
head(healthcare_data)

```
\newpage
## HCAI Healthcare dataset con't
```{r}
#change "." to "_" for consistency with other datasets
names(healthcare_data) <- gsub(x = names(healthcare_data), pattern = "\\.", 
                               replacement = "_")  
head(healthcare_data)

#change county names to match other two datasets (remove numbers in front)
healthcare_data<-healthcare_data%>%
  mutate(county=substring(healthcare_data$county, 6))
head(healthcare_data)
```
\newpage
## HCAI Healthcare dataset con't
```{r}
#change money column ("Total.Costs...") to numeric
healthcare_data<-healthcare_data%>%
  mutate(total_costs_of_oshpd_projects = total_costs_of_oshpd_projects %>% 
           str_remove_all("[$,]"))
healthcare_data$total_costs_of_oshpd_projects<-as.numeric(healthcare_data$total_costs_of_oshpd_projects)
head(healthcare_data)

#fix "data_generation_date" (remove empty time)
healthcare_data$data_generation_date<-as.Date(healthcare_data$data_generation_date)
head(healthcare_data)

```
\newpage

# Description of dataset
## What is the data source? How does the dataset relate to the group problem 
## statement and question?
### 1. Demographic: Census Data
Includes 58 observations with 22 variables. All are numeric or integers except
for the first column which is a character with county names.

### 2. Mortality: CA open data portal, California Department of Public Health 
Includes 147,784 observations with 10 variables. most are characters except for 
year and count of mortality.

### 3. Healthcare: CA open data portal, Department of Healthcare Access and Information
Includes 53,592 observations with 6 variables. 

#### How does the dataset relate to the group problem statement and question?

# Identify data types for 5+ data elements/columns/variables
### 1. 
### 2. 
### 3. 
### 4. 
### 5.   
### 6.   
### 7.   

## Utilize functions or resources in RStudio to determine the types of each data 
## element (i.e. character, numeric, factor)
### 1. 
### 2. 
### 3. 
### 4. 
### 5. 
### 6.   
### 7.   
\newpage

## Identify the desired type/format for each variable—will you need to convert 
## any columns to numeric or another type? 
### 1. 
### 2. 
### 3. 
### 4. 
### 5.   
### 6.   
### 7.   

## Provide a basic description of the 5+ data elements (Numeric: mean, median, 
## range; Character: unique values/categories)
### 1. 
### 2. 
### 3. 
### 4. 
### 5. 
### 6.   
### 7.   


